Motor Learning Without Moving: Proprioceptive and Predictive Hand Localization

### 2019-04-26 paper/paper.Rmd
revisions of paper

---
title: "Motor Learning Without Moving: Proprioceptive and Predictive Hand Localization After Passive Visuoproprioceptive Discrepancy Training"
author:
  - name: Ahmed A. Mostafa
    affiliation: [YORK,MANSOURA]
#    affiliation: Faculty of Physical Education, Mansoura University, Mansoura, Egypt
  - name: Bernard Marius 't Hart
    affiliation: YORK
    email: thartbm@gmail.com
    corresponding: thartbm@gmail.com
  - name: Denise Y.P. Henriques
    email: deniseh@yorku.ca
    affiliation: YORK
address:
  - code: YORK
    address: CVR / Kinesiology and Health Science, York University, Toronto, Ontario, Canada
  - code: MANSOURA
    address: Faculty of Physical Education, Mansoura University, Mansoura, Egypt

abstract: |
  An accurate estimate of limb position is necessary for movement planning. Where we localize our unseen hand after a reach depends on felt hand position, or proprioception, but this is usually neglected in favour of predicted sensory consequences based on efference copies of motor commands. Both sources of information should contribute, so here we set out to further investigate how much of hand localization depends on proprioception and how much on predicted sensory consequences. We use a passive training paradigm with rotated visual feedback that eliminates the possibility to update predicted sensory consequences, but still recalibrates proprioception. After this training we measure participants’ hand location estimates based on both efference-based predictions and afferent proprioceptive signals with self-generated hand movements as well as based on proprioception only with robot-generated movements. We find indistinguishable shifts in hand localization after robot- and self-generated hand movements, and changes in open-loop reaches. Both motor and proprioceptive changes are only slightly smaller as those after training with self-generated movements, confirming that proprioception plays a large role in estimating limb position and in planning movements. (data: https://doi.org/10.17605/osf.io/zfdth, preprint: https://doi.org/10.1101/384941)

header-includes:
   - \usepackage{threeparttable}
   - \usepackage[table]{xcolor}
   - \usepackage{graphicx}
   - \usepackage{tabularx}
   - \usepackage{longtable}
   - \usepackage{siunitx}
bibliography: bibliography.bib
output: rticles::plos_article
csl: plos.csl
---

@@

Sensory information is central to how we control all our movements. Our brain is even thought to use predicted sensory consequences derived from efferent copies of motor commands for motor control @Duhamel1992a. When training with rotated visual feedback of the hand, we update these predictions @Wolpert1995. Additionally, such training leads to a recalibration of our sense of felt hand position - “proprioception” - to be more aligned with the distorted visual feedback @Cressman2009. Both of these changes have been measured by asking people to localize where their unseen hand is – before and after training [@Clayton2013; @Izawa2012b; @Jones2012a; @Synofzik2008; @Yavari2016]. While our lab has previously found that proprioception accounts for a large portion of the change in hand localization @Hart2016, it is far from clear how much each process contributes or how to tease them apart. Here we use passive training, that removes the need to update predicted sensory consequences, in an attempt to isolate the contribution of proprioception to hand localization.

The predicted sensory consequences of movements may play several important roles in motor learning and control. Predicted sensory consequences allow us to correct our movements before sensory error signals are available, they can be used to select movements that best achieve our goals and they may inform us on the location of our limbs. Hence measuring predicted sensory consequences is valuable in movement research. In visuomotor rotation adaptation tasks, the actual sensory outcome is systematically different from the expected outcome, so that participants update their predictions on the outcome of the trained movements. In previous experiments, people were asked to make a movement and then indicate the location of, or “localize,” their unseen, right hand, before and after training with rotated visual feedback [@Izawa2012b; @Synofzik2008]. Since there was no visual information available to the participants, the predicted sensory consequences of the movement should be used in localizing the unseen hand. Both studies found a significant shift in hand localization, providing evidence that predicted sensory consequences are indeed updated as a result of visuomotor rotation adaptation.

However, our lab has shown that our sense of where we feel our hand to be, proprioception, is also reliably recalibrated after visuomotor rotation adaptation [@Cressman2009; @Hart2016; @Mostafa2014; @Mostafa2015; @Maksimovic2018; @Henriques2012; @Cressman2010a; @Barkley2014; @Salomonczyk2011; @Nourouzpour2014; @Henriques2014]. This has also been shown by other labs @Cameron2012 and a comparable proprioceptive change is induced with force-field adaptation @Ostry2016. As proprioception also informs us on the location of our limbs, we have on occasion used a task that is very similar to hand localization to investigate this [@Cressman2009; @Jones2012a; @Clayton2014; @Ruttle2018]. Although proprioceptive recalibration has been largely ignored as an explanation for changes in hand localization, we and others have shown that it accounts for a substantial part of the changes in localization, along with updates in predicted sensory consequences [@Hart2016; @Cameron2012]. Nevertheless, it is far from clear how much proprioception and prediction each contribute to hand localization.

Here we intend to further examine the contribution of proprioception to hand localization. To do this, we use passive training, where a robot arm moves the participant’s arm out, so that the cursor always directly goes to the target [@Cameron2012; @Cressman2010; @Salomonczyk2013]. This means there is no efference copy available and no visuomotor error-signal, both of which are required to update predicted sensory consequences. However, we impose a discrepancy between vision and proprioception that drives proprioceptive recalibration. Thus, changes in hand localization after this type of training should be due to proprioceptive recalibration only. We use the same experimental protocol as before @Hart2016, so that we can compare localization shifts between the two different types of training, and can better assess the contributions of predicted sensory consequences and proprioception to hand localization.

@@

We set out to test the relative contributions of proprioception and efference-based prediction to hand localization. We use visual training with robot-generated hand movements to prevent updates of predicted sensory consequences, but still elicit proprioceptive recalibration.

@@

Twenty-five right-handed participants were recruited for the passive learning task in this study. One participant was excluded for not following task instructions, and three were excluded for low performance on a task that ensures attention during the passive training. All analyses presented here pertain to the remaining twenty-one participants (13 females and 8 males, mean age: 20.1 +/- 2.3 years), but the data of the three low-performing participants is included in the online \href{https://osf.io/zfdth}{dataset} @Mostafa2017. Participants provided prior, written informed consent in accordance with the ethical guidelines set by the York Human subjects Review Subcommittee (#2014-240). All had normal or corrected-to-normal vision, and received credit toward an undergraduate psychology course. Participants were screened verbally, and all reported being right handed and not having any history of visual, neurological, and/or motor dysfunction.  Published data from 21 further participants is used for comparison @Hart2016.

![**Setup and tasks:** **a)** Participants moved their unseen right hand with visual feedback on hand position provided through a mirror (middle surface) half-way between their hand and the monitor (top surface). A touchscreen located just above the hand was used to collect responses for the localization tasks (bottom surface). **b)** Training task. The target, shown as a yellow disc, is located 10 cm away from the home position at 45°. In the rotated training tasks, the cursor (shown here as a green circle) represents the hand position rotated 30° relative to the home position. **c)** No-cursor reach task. Targets are located 10 cm away from the home position at 15°, 25°, 35°, 45°, 55°, 65°, and 75°, shown by the yellow circles here (only one was shown on each trial). While reaching to one of these targets, no visual feedback on hand position is provided. d) Localization task. The participants’ unseen, right hands moved out, and subsequently participants indicated the direction of the hand movement by indicating a location on an arc using a touch screen with their visible left index finger](Fig1_setup.eps)

@@

Participants sat in a height-adjustable chair to ensure that they could easily see and reach all targets presented on a reflective screen (see Fig 1). During all tasks, they held the vertical handle on a two-joint robot manipulandum (Interactive Motion Technologies Inc., Cambridge, MA) with their right hand so that their thumb rested on top of the handle. A monitor (Samsung 510 N, refresh rate 60 Hz) was mounted 11 cm above the reflective screen, such that images displayed on the monitor appeared to lie in the horizontal plane where the right hand was moving. The reflective screen was mounted horizontally 18 cm above the robot manipulandum. A touch screen was mounted 13 cm underneath the reflective surface, so that subjects could indicate the location of the unseen right-hand locations (specifically the unseen thumb) with their visible left hand, which was lit up with a small spot light (only in localization tasks). The room lights were dimmed and the participants’ view of their right hand was blocked by the reflective screen, as well as a dark cloth draped between the touch screen and participants’ right shoulders.

@@

The first part of the experiment used training with a cursor aligned with the hand and the second part had training with a cursor rotated around the start position (Fig 1b; white rows in Table 1). During the training with rotated feedback, the cursor was gradually rotated 30° clockwise. This introduced a discrepancy between the actual, or felt, hand position and the visual feedback, that should evoke proprioceptive recalibration. However, the movements were robot generated, so that there were no predicted sensory consequences based on the outgoing motor command. Hence the prediction errors that are thought to lead to motor learning were absent as well. After both types of training, participants did open-loop reaches as well as two kinds of hand localization tasks, to test the effect of training on proprioceptive and predictive hand estimates.

\newcommand{\headrow}{\rowcolor{black!20}}
\newcommand{\thead}[1]{\multicolumn{1}{l}{\bfseries #1\rule[-1.2ex]{0pt}{2em}}}

\definecolor{LGray}{gray}{0.9}
\definecolor{DGray}{gray}{0.8}

\begin{table}[bt]
\caption{Task order}
\begin{threeparttable}
\begin{tabular}{cccc}
\headrow
\multicolumn{2}{c}{\# trials} & \multicolumn{2}{c}{training type} \\
\headrow
%\multicolumn{2}{|c|}{\# trials} & \thead{Exposure training} & \thead{Classic training} \\
\thead{aligned} & \thead{rotated} & \multicolumn{1}{c}{\textbf{Exposure training}} & \multicolumn{1}{c}{\textbf{Classic training}} \\
\rowcolor{white}
50 & 90 & training & training \\
\rowcolor{LGray}
- & 21 & no-cursor & no-cursor \\
\rowcolor{white}
- & 60 & training & training \\
\rowcolor{DGray}
25 & 25 & active delayed localization & active delayed localization \\
\rowcolor{LGray}
21 & 21 & no-cursor & no-cursor \\
\rowcolor{white}
10 & 60 & training & training \\
\rowcolor{DGray}
25 & 25 & passive delayed localization & passive delayed localization \\
\rowcolor{LGray}
21 & 21 & no-cursor & no-cursor \\
\rowcolor{white}
10 & 60 & training & training \\
\rowcolor{DGray}
25 & 25 & active delayed localization & \textit{(active online localization)} \\
\rowcolor{LGray}
21 & 21 & no-cursor & no-cursor \\
\rowcolor{white}
10 & 60 & training & training \\
\rowcolor{DGray}
25 & 25 & passive delayed localization & \textit{(passive online localization)} \\
\rowcolor{LGray}
21 & 21 & no-cursor & no-cursor \\
\hiderowcolors
\hline  % Please only put a hline at the end of the table
\end{tabular}

\begin{tablenotes}
\item 
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

In what we call ‘exposure training’ the participants did not move their hand toward the target, but the robot did. In this task (Table 1), the right hand was represented by a cursor (green disk, 1 cm in diameter, Fig 1b) located directly above participant’s thumb. The robot moved the participant’s unseen right hand (and the cursor) along a direct path toward a visual yellow target disk and back to the starting position (1 cm in diameter, Fig 1b). The home position was located approximately 20 cm in front of participants and the visual target located 10 cm from the home position at 45° (Fig 1b). In order to make sure participants were paying attention to the cursor, the cursor was switched off for 2 screen refreshes (~33.3ms) on 50\% of the trials at a random distance between 4 and 9 cm from the home position and participants were asked to report this using a button press with the left hand. Performance on this task was used to screen participants.

During the first half of the experiment, the cursor and hand path were aligned during exposure training. In the second part of the experiment, the “rotated” session, the same visual training target at 45° was used, and the cursor kept moving straight to this target. However, the robot-generated hand path gradually rotated 30° CCW (Fig 1b) with respect to the visible target and the cursor in increments of 0.75°/trial, so that the full rotation was reached after 45 trials. This mimicked error-free responses to a gradual visuomotor rotation of 30° CW. The initial training consisted of 50 trials in the aligned part and 90 in the rotated part. In between open-loop reach tasks and localization tasks (Table 1) extra training tasks were done, each of which consisted of 10 trials in the aligned part of the experiment and 60 trials in the rotated part (Table 1).

@@

The trials in no-cursor reaching (Table 1, light gray rows) serve as a classical measure of motor adaptation. On each of these trials participants were asked to reach with their unseen right hand to one of 7 visual targets, without any visual feedback of hand position. The targets were 10 cm from the home position, located radially at: 15°, 25°, 35°, 45°, 55°, 65°, and 75° (Fig 1c). A trial started with the robot handle at the home position and, after 500 ms, the home position disappeared and the target appeared, cuing the participants to reach for the target. Once the participants thought they had reached the target they held their position for 250 ms, and the target and the home position disappeared, cuing participants to move back to the home position along a straight, constrained path, to begin the next trial. If participants tried to move outside of the path, a resistance force, with a stiffness of 2 N/(mm/s) and a viscous damping of 5 N/(mm/s), was generated perpendicular to the path. In every iteration of the no-cursor reach task, each target was reached three times, for a total of 21 reaches in pseudo-random order. The no-cursor reaching task was performed four times in the aligned part of the experiment and five times in the rotated part of the experiment.

@@

In this task (Fig 1d; Table 1, dark gray rows) participants indicated where they thought their unseen right hand was after a movement. First, an arc appeared, spanning from 0° to 90° and located 10 cm away from the home position and the participants’ unseen, right hand moved out from the home position in a direction towards towards a point on the arc. The hand was stopped by the robot at 10 cm from the home position and then, to prevent the online proprioceptive signals from overriding the predictive signals [@Izawa2012b; @Hart2016], the hand was moved back to the home position using the same kind of constrained path as used for the return movements in the no-cursor task. Participants indicated with the index finger of their visible, left hand on the touch screen mounted directly above the robot handle where they thought their trained hand had crossed the arc.

Crucially, there were two variations of this task. First, in the ‘active’ localization task participants generated the movement themselves, as they could freely move their unseen right hand from the home position to any point on the arc. Second, there was a ‘passive’ localization task where the robot moved the participants’ hand out and back, to the same locations the participants moved to in the preceding ‘active’ localization task in a shuffled order (hence, active localization is done first). In active localization, participants have access to both proprioceptive information as well as an efference-based prediction of sensory consequences, but in passive localization, only proprioception should be available. The active and passive localization task each consisted of 25 trials, and each of the tasks was done a total of four times; twice after aligned and twice after rotated training.

@@

The paradigm described above is an exact replica of a paradigm we used earlier @Hart2016 with two exceptions. First, we used exposure training here, instead of the standard reach training with volitional movements, which we will call ‘classic’ training. Second, all localization is delayed until the right hand has returned to the home position in this study (see Table 1), so that instead of both delayed and online localization we have two repetitions of each delayed localization task. With this paradigm we can compare changes in localization and no-cursor reaches change after exposure training with changes in the same measures after classic visuomotor adaptation training.

@@

Localization responses were taken as the (signed) angular difference between vectors through the home position and the actual hand position as well as the location indicated on the touch screen. Prior to analyses, idiosyncratic differences in performing this task were countered. Before conversion to degrees angle, a circle with a 10 cm radius was fit to the touch screen responses of each participant and the offset of this circle’s centre was subtracted from all response coordinates, so that all responses fell close to the arc. Then, a smoothed spline was fit to every participant’s response errors in each of the four localization tasks (aligned vs. rotated and active vs. passive) and these were used to obtain localization errors at the same locations used for the no-cursor reaches (15°, 25°, 35°, 45°, 55°, 65° and 75°), but only if that location fell within the range of the data (i.e. we only interpolate). This way localization responses could be compared across participants despite the freely chosen reach directions. At the 15° location 7/21 participants didn’t have an estimate in one or more of the four localization tasks (in the “classic” data it was also 7/21). While that data is shown in the figures, we did not use it for analysis.

First we test if localization responses shifted following rotated exposure training compared to aligned. We then test if the shift in localization responses is different for active and passive localization, and we run analyses comparing localization after exposure training with localization after “classic” training. Finally, we explore the generalization of localization responses and if they are different between the groups doing classic and exposure training.

@@

To assess any reach adaptation that may have occurred after exposure training we analyzed reach endpoint errors in no-cursor trials. Reach endpoint errors were the (signed) angular difference between a vector from the home position to reach endpoint and a vector from the home position to the target. We obtained reach aftereffects by subtracting reach endpoint errors after aligned training from those after rotated training. No-cursor endpoint errors were analyzed to see if participants adapted the direction of their reaching movements after rotated exposure training. We also tested if any such change decayed, i.e. if it was the same immediately after exposure training, or when a localization task was done in between exposure training and no-cursor reaches. Furthermore, we tested if the generalization of reach aftereffects is different between exposure and classic training and if there is any generalization of reach aftereffects.

>>>>

---
title: 'Motor Learning Without Moving: Proprioceptive and Predictive Hand Localization
  After Passive Visuoproprioceptive Discrepancy Training'
author:
- affiliation:
  - YORK
  - MANSOURA
  name: Ahmed A. Mostafa
- affiliation: YORK
  corresponding: thartbm@gmail.com
  email: thartbm@gmail.com
  name: Bernard Marius 't Hart
- affiliation: YORK
  email: deniseh@yorku.ca
  name: Denise Y.P. Henriques
address:
- address: CVR / Kinesiology and Health Science, York University, Toronto, Ontario,
    Canada
  code: YORK
- address: Faculty of Physical Education, Mansoura University, Mansoura, Egypt
  code: MANSOURA
output:
  html_document:
    df_print: paged
  pdf_document: default
  rticles::plos_article: default
  word_document: default
csl: plos.csl
fig_caption: yes
header-includes:
- \usepackage{threeparttable}
- \usepackage[table]{xcolor}
- \usepackage{graphicx}
- \usepackage{tabularx}
- \usepackage{longtable}
- \usepackage{siunitx}
bibliography: bibliography.bib
abstract: |
  An accurate estimate of limb position is necessary for movement planning, before and after motor learning. Where we localize our unseen hand after a reach depends on felt hand position, or proprioception, but in studies and theories on motor adaptation this is quite often neglected in favour of predicted sensory consequences based on efference copies of motor commands. Both sources of information should contribute, so here we set out to further investigate how much of hand localization depends on proprioception and how much on predicted sensory consequences. We use a training paradigm combining robot controlled hand movements with rotated visual feedback that eliminates the possibility to update predicted sensory consequences (‘exposure training’), but still recalibrates proprioception, as well as a classic training paradigm with self-generated movements in another set of participants. After each kind of training we measure participants’ hand location estimates based on both efference-based predictions and afferent proprioceptive signals with self-generated hand movements (‘active localization’) as well as based on proprioception only with robot-generated movements (‘passive localization’). In the exposure training group, we find indistinguishable shifts in passive and active hand localization, but after classic training, active localization shifts more than passive, indicating a contribution from updated predicted sensory consequences. Both changes in open-loop reaches and hand localization are only slightly smaller after exposure training as compared to after classic training, confirming that proprioception plays a large role in estimating limb position and in planning movements, even after adaptation. (data: https://doi.org/10.17605/osf.io/zfdth, preprint: https://doi.org/10.1101/384941)
---

@@

Sensory information is central to how we control all our movements. Our brain is even thought to use predicted sensory consequences derived from efference copies of motor commands for motor control @Duhamel1992a. When training with rotated visual feedback of the hand, we update these predictions @Wolpert1995, and these updated predictions contribute to motor control. Additionally, such training leads to a recalibration of our sense of felt hand position based on afferent signals (that we call “proprioception” here) to be more aligned with the distorted visual feedback @Cressman2009. Both of these changes have been measured by asking people to localize where their unseen hand is – before and after training [@Clayton2013; @Izawa2012b; @Jones2012a; @Synofzik2008; @Yavari2016]. While our lab has previously found that proprioception accounts for a substantial portion of the change in hand localization, and should hence contribute to motor adaptation @Hart2016, it is far from clear how much each process contributes or how to tease them apart. Here we use training with robot-generated reaches, that prevents efference-based predictions and eliminates differences between actual and expected visual feedback and hence prevents updating predicted sensory consequences. We then compare hand localization data from this experiment to hand localization in a classic training paradigm with self-generated movements, from an – otherwise identical – earlier study @Hart2016, in an attempt to isolate the contribution of proprioception to hand localization prior to and after adaptation.

The predicted sensory consequences of movements likely play several important roles in motor learning and control. Predicted sensory consequences allow us to correct our movements before sensory error signals are available, they can be used to select movements that best achieve our goals and they may inform us on the location of our limbs. Hence measuring predicted sensory consequences is valuable in movement research. In motor adaptation, the actual sensory outcome is systematically different from the expected outcome, so that participants update their predictions on the outcome of the trained movements. In two previous visuomotor adaptation experiments, people were asked to make a movement and then indicate the location of, or “localize,” their unseen, right hand, before and after training with rotated visual feedback [@Izawa2012b; @Synofzik2008]. Since there was no visual information available to the participants, the predicted sensory consequences of the movement should be used in localizing the unseen hand. Both studies found a significant shift in hand localization, providing evidence that predicted sensory consequences are indeed updated as a result of visuomotor rotation adaptation.

However, there is another explanation for shifts in hand localization after visuomotor adaptation. Our lab, and others, have shown that our sense of where we feel our hand to be, proprioception, is also reliably recalibrated after visuomotor adaptation [@Cressman2009; @Hart2016; @Mostafa2014; @Mostafa2015; @Maksimovic2018; @Henriques2012; @Cressman2010a; @Barkley2014; @Salomonczyk2011; @Nourouzpour2014; @Henriques2014; @Cameron2012], and a comparable proprioceptive change is induced with force-field adaptation [@Ostry2016; @Ohashi2019]. As proprioception, like predicted sensory consequences, also informs us on the location of our limbs, we have on occasion used a task that is very similar to hand localization to investigate this [@Cressman2009; @Jones2012a; @Clayton2014; @Ruttle2018]. Although proprioceptive recalibration has been largely ignored as an explanation for changes in hand localization, we and others have shown that it accounts for a substantial part of the changes in localization, along with, but separate from, updates in predicted sensory consequences [@Hart2016; @Cameron2012]. Nevertheless, it is far from clear how much proprioception and prediction each contribute to hand localization.

Here we intend to further examine the contribution of proprioception to hand localization, relative to the contribution of predicted sensory consequences. To do this, we use a passive training paradigm with robot-generated  movements, which we refer to as ‘exposure training’, where the cursor always goes directly goes to the target [@Cameron2012; @Cressman2010; @Salomonczyk2013]. This means there is no efference copy available and no visuomotor error-signal, both of which are required to update predicted sensory consequences. That is, exposure training to visuoproprioceptive discrepancy should not lead to updates of predicted sensory consequences. However, the discrepancy between vision and proprioception does drive proprioceptive recalibration. Thus, changes in any kind of hand localization after this type of training should be due to proprioceptive recalibration only. We use the same experimental protocol as before @Hart2016, where classic visuomotor training should have changed predicted sensory consequences along with proprioception. This allows comparing shifts in hand localization between the two different types of training, and a better assessment of the contributions of predicted sensory consequences and proprioception to hand localization.

Primarily, we will compare the magnitude of shifts in passive and active hand localization after visuoproprioceptive discrepancy exposure and classic visuomotor training. After exposure training there should be no difference, while after classic training, predicted sensory consequences should be updated but only contribute to active localization, so that active and passive localization changes should be different. For a secondary analysis we will compare the magnitude of reach aftereffects between exposure and classic training. As exposure training should not lead to updates of internal models, exposure training should lead to lower reach aftereffects than classic training. However, if there are any reach aftereffects at all after exposure training this means that afferent-based localization shifts drive motor changes too, and here we can see to what degree that is the case.

@@

We set out to test the relative contributions of proprioception and efference-based prediction to hand localization. We use exposure training, with robot-generated reaches to prevent updates of predicted sensory consequences, but still elicit proprioceptive recalibration, and compare this with a classic training paradigm that should change both predicted sensory consequences and proprioception. We then tested this in both active localization, that should use both sources – updated or not – and in passive localization, that should rely only on proprioception, which should be recalibrated equally by both training paradigms.

@@

Twenty-five right-handed participants were recruited for the exposure training group in this study. One participant was excluded for not following task instructions, and three were excluded for low performance on a task that ensures attention during the exposure training. All analyses presented here pertain to the remaining twenty-one participants (mean age: 20.1±2.3, age range: 18-25, 13 females), but the data of the three low-performing participants is included in the online dataset @Mostafa2017 ([https://osf.io/zfdth](https://osf.io/zfdth/)). Participants provided prior, written informed consent in accordance with the ethical guidelines set by the York Human subjects Review Subcommittee (#2014-240). All had normal or corrected-to-normal vision, and received credit toward an undergraduate psychology course. Participants were screened verbally, and all reported being right handed and not having any history of visual, neurological, and/or motor dysfunction. Published data from 21 further participants (mean age: 24, age range: 18-38, 11 females) that completed classic visuomotor training is used for comparison @Hart2016.

![**Setup and tasks** **a)** The participants held on to the handle of a robotic manipulandum with their unseen right hand. Visual feedback on hand position was provided through a mirror (_middle surface_) half-way between the hand and the monitor (_top surface_). A touchscreen located just above the hand was used to collect responses for the localization tasks (_bottom surface_). **b)** Training task. The target, shown as a yellow disc, is located 10 cm away from the home position at 45°. In the rotated training tasks, the cursor (shown here as a green circle) represents the hand position rotated 30° relative to the home position. **c)** No-cursor reach task. Targets are located 10 cm away from the home position at 15°, 25°, 35°, 45°, 55°, 65°, and 75°, shown by the yellow circles here (only one was shown on each trial). While reaching to one of these targets, no visual feedback on hand position is provided. **d)** Localization task. The participants’ unseen, right hands moved out and back, and subsequently participants indicated the direction of the hand movement by indicating a location on an arc using a touch screen with their visible left index finger.](Fig1_setup.eps)

@@

Participants sat in a height-adjustable chair to ensure that they could easily see and reach all targets. To ensure no visual information on right hand position was available to localize the right hand, the room lights were dimmed and the participants’ view of their right hand was blocked by a horizontal surface, as well as a dark cloth draped between a touch screen and participants’ right shoulders. During all tasks, they held the vertical handle on a two-joint robot manipulandum (Interactive Motion Technologies Inc., Cambridge, MA) with their right hand so that their thumb rested on top of the handle. A horizontal touch screen, reflective surface and monitor were mounted 5, 18 and 29 cm above the handle, respectively. Stimuli were presented using the monitor (Samsung 510 N, refresh rate 60 Hz), which was mounted facing downward, 11 cm above the reflective surface, such that images displayed on the monitor and viewed in the reflective surface appeared to lie in the horizontal plane where the right hand was moving. The reflective surface was mounted horizontally 18 cm above the robot manipulandum. The touch screen was mounted 13 cm underneath the reflective surface (5 cm above the manipulandum), so that subjects could indicate the location of the unseen right hand (specifically the thumb on top of the handle) with their left hand. The touch screen always blocked vision of the right hand, but the left hand was lit up with a small spot light (only in localization tasks) in order to be visible through the reflective surface for accurate touch screen responses. For a schematic illustration of the setup, see Fig 1a.

@@

The first part of the experiment used training with a cursor aligned with the hand and the second part had training with a cursor rotated around the start position (Fig 1b; Fig 2a). During the training with rotated feedback, the cursor was gradually rotated 30° clockwise. This introduced a discrepancy between the actual, or felt, hand position and the visual feedback, that should evoke proprioceptive recalibration. However, the movements were robot generated, and always brought the cursor straight to the target, so that there were no motor commands generated, and no predicted sensory consequences based on this outgoing command. In the absence of predicted sensory consequences and with perfect, error-free reach performance, there were no sensory prediction errors to drive classic visuomotor adaptation either @Wong2019. After both types of training, participants did two kinds of hand localization tasks, to test the effect of training on proprioceptive and predictive hand estimates, as well as open-loop reaches as a classic measure of implicit adaptation.

![**Experimental design** **a)** Task order. All participants performed active and passive localization as well as no-cursor reaching tasks in both a first, aligned session and a second, rotated session. There is an extra no-cursor reach block in the rotated session, to assess decay. The number of trials (numbers above the rectangles) in the no-cursor reaching tasks and both types of localization were the same in the aligned and rotated sessions, but while aligned training was kept to a minimum, rotated training was increased to ensure sufficient adaptation. The Exposure training group did passive and active localization twice, but the Classic training group did different localization tasks in the second half of each session, the data of which is not used here. **b)** Our main question was about the signals contributing to hand localization in the four different kinds of localization. According to our hypothesis, updated predicted sensory consequences should have only contributed to active hand localization after Classic training.](Fig2.eps)

@@

In what we called ‘exposure training’ the participants did not move their hand toward the target, but the robot did. In this task (Fig 2a, gray boxes), the right hand was represented by a cursor (green disk, 1 cm in diameter, Fig 1b) located directly above participant’s thumb. The robot moved the participant’s unseen right hand (and the cursor) along a direct path, with a constant velocity of 10 cm/s, toward a visual yellow target disk and back to the starting position (1 cm in diameter, Fig 1b). The home position was located approximately 20 cm in front of participants and the visual target located 10 cm from the home position at 45° (Fig 1b). In order to make sure participants were paying attention to the cursor, the cursor was switched off for 2 screen refreshes (~33.3ms) on 50\% of the trials at a random distance between 4 and 9 cm from the home position and participants were asked to report this using a button press with the left hand. Performance on this task was used to screen participants.

During the first half of the experiment, the “aligned” session, the cursor and hand path were aligned during exposure training. In the second part of the experiment, the “rotated” session, the visuoproprioceptive discrepancy was introduced. The same visual training target at 45° was used, and the cursor kept moving straight to this target. However, the robot-generated hand path gradually rotated 30° CCW (Fig 1b) with respect to the cursor and target, in increments of 0.75°/trial, to prevent awareness of the rotation, so that the full rotation was reached after 45 trials. This mimicked error-free responses to a gradual visuomotor rotation of 30° CW. The initial training consisted of 50 trials in the aligned part and 90 in the rotated part. In between open-loop reach tasks and localization tasks (Fig 2, outlined and blue boxes) extra training tasks were done, each of which consisted of 10 trials in the aligned part of the experiment and 60 trials in the rotated part (Fig 2, gray boxes).

@@

The trials in no-cursor reaching (Fig 1c; Fig 2, outlined boxes) served as a standard measure of motor adaptation. On each of these trials participants were asked to reach with their unseen right hand to one of 7 visual targets, without any visual feedback of hand position. The targets were 10 cm from the home position, located radially at: 15°, 25°, 35°, 45°, 55°, 65°, and 75° (Fig 1c). A trial started with the robot handle at the home position and, after 500 ms, the home position disappeared and the target appeared, cuing the participants to reach for the target. Once the participants thought they had reached the target they held their position for 250 ms, and the target disappeared, while the home position re-appeared, cuing participants to move back to the home position along a straight, constrained path, to begin the next trial. The path was constrained so that if participants tried to move outside of the path, a resistance force, with a stiffness of 2 N/(mm/s) and a viscous damping of 5 N/(mm/s), was generated perpendicular to the path. In every iteration of the no-cursor reach task, each target was reached for three times, for a total of 21 reaches in pseudo-random order. The no-cursor reaching task was performed four times in the aligned part of the experiment and five times in the rotated part of the experiment. The end-points of no-cursor reaches, in particular the angles relative to the start position and their deviations from the angles of the targets were analyzed, as these typically change in classic visuomotor adaptation.

@@

In this task (Fig 1d; Fig 2, blue boxes) participants indicated where they thought their unseen right hand was after a movement. First, an arc appeared, spanning from 0° to 90° and located 10 cm away from the home position and the participants’ unseen, right hand moved out from the home position in a direction towards a point on the arc. The hand was stopped by the robot at 10 cm from the home position and then, to prevent online proprioceptive signals from overriding the predictive signals [@Izawa2012b; @Hart2016], the hand was moved back to the home position using the same kind of constrained path as used for the return movements in the no-cursor task. Participants indicated with the index finger of their visible, left hand on the touch screen mounted directly above the robot handle where they thought their trained hand had ended its movement under the arc.

Crucially, there were two variations of this task. First, in the ‘active’ localization task participants generated the movement themselves, as they could freely move their unseen right hand from the home position to any point on the arc (Fig 2a, dark blue boxes). Second, there was a ‘passive’ localization task where the robot moved the participants’ hand out and back (Fig 2a, light blue boxes), to the same locations the participants moved to in the preceding ‘active’ localization task in a shuffled order. Hence, active localization was always done first. In active localization, participants had access to both afferent, proprioceptive information as well as an efference-based prediction of sensory consequences, but in passive localization, only proprioception should have been available. The active and passive localization task each consisted of 25 trials, and each of the tasks was done a total of four times; twice after aligned and twice after rotated training. From these tasks, the localization responses were analyzed, i.e. where participants indicated the endpoint of their hand movement was, in particular the angles of the indicated points relative to the home position in comparison to the true reach endpoints’ angle.

@@

The paradigm described above was an exact replica of a paradigm we used earlier @Hart2016 with three exceptions. First, we used exposure training here, instead of the standard reach training with volitional movements, which we called ‘classic’ training. Second, all localization was delayed until the right hand has returned to the home position in this study, whereas the previous study also included a second set of “online” localization tasks, where touch screen responses were collected when the unseen, right hand was still at the furthest point of the movement. Instead of both delayed and online localization we had two repetitions of each delayed localization task here. Third, we added cursor blink detection to the training tasks, to make sure participants attended to the task. In the present study we compared changes in localization responses and no-cursor reach endpoints after exposure training with changes in the same measures after classic visuomotor adaptation training, as collected earlier.

@@

We primarily wanted to test if hand localization shifts after exposure training, whether or not there were differences between active and passive localization shifts, and finally compare localization shifts after exposure training to those after classic visuomotor training.

There were some idiosyncratic differences in performing localization (e.g. systematic under- or overshoot of the arc by some participants). To counter this, we fitted a circle with a 10 cm radius to the touch screen responses of each participant (using a least-squared errors algorithm) and the offset of this circle’s centre was subtracted from all response coordinates, so that all responses fell close to the arc. Localization responses were taken as the (signed) angular difference between vectors through the home position and both the actual hand position as well as the location indicated on the touch screen. For each localization response, we retain the angle of the actual reach endpoint, as well as the signed angular error of the localization response. Since participants freely chose their movement directions, we interpolated angular localization errors at the same angles used for the no-cursor reaches (15°, 25°, 35°, 45°, 55°, 65° and 75°), but only if that location fell within the range of the data (i.e. we did not extrapolate). For interpolation we used a smoothed spline that was fit to every participant’s response errors in each of the four localization tasks (aligned vs. rotated and active vs. passive). This way localization responses could be compared across participants despite the freely chosen reach directions. At the 15° location 7/21 participants didn’t have an estimate in one or more of the four localization tasks (in the “classic” data it was also 7/21). While that data was shown in the figures, we did not use it for analysis.

First we compared localization responses shifted following rotated exposure training to those following aligned, and confirmed that training shifted localization. We then tested if the shift in localization responses is different for active and passive localization, and we ran analyses comparing localization after exposure training with localization after “classic” training. Finally, we test if the variance of active localization responses is lower than the variance of passive localization responses.

@@

For our secondary question, we wanted to assess if any reach adaptation had occurred after exposure training, so we analyzed reach endpoint errors in no-cursor trials. Reach endpoint errors were the (signed) angular difference between a vector from the home position to reach endpoint and a vector from the home position to the target. We obtained reach aftereffects by subtracting median reach endpoint errors for each of the 7 targets after aligned training from those after rotated training. No-cursor endpoint errors were analyzed to test if participants adapted the direction of their reaching movements after rotated exposure training, which would indicate that recalibrated proprioception contributes to motor changes. We also tested if any such change decayed, i.e. if it was the same immediately after exposure training, or when a localization task was done in between exposure training and no-cursor reaches. Finally, we compare reach aftereffects between the exposure and classic training groups.

## Generalization maxima
